Iterative Monomorphisation

### 2025-02-21 iterative_monomorphisation.tex
Jannisg

\begin{examplex}\rm
Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] 
         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

\looseness=-1
The algorithm presented above is too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This dispenses from having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the pseudocode are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between type argument tuples having already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, we keep track of a map \(S\) from formulae to the substitutions generated by the matchings. It is used to generate new formulae in the final phase.

All sets referenced in the pseudocode are %assumed to be
finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, will match \(\upsilon\) against \(\tau\) and either fail or return \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching. The pseudocode matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
   returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} %, where \(\sigma_1\) and \(\sigma_2\) are substitutions,
   tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
%   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae, returns the same set of formulae where all types have been mangled.
\end{enumerate}

@@

   \Return \(\Mangle(\GenFormulae(\Phi, S))\)
}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for iterative monomorphisation}
\label{iter_mono}
\end{figure}

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no general support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

@@

   \ForEach{\(\sigma\in \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
   \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))\)}{
   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
            \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\)}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\)\;
                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }{\textcolor{ourblueviolet}{
               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\)\;
               \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }}
            }}
      }
   }

@@

\end{algorithm}
%\end{quote}
\caption{Pseudocode for formula monomorphisation step}
\label{mono_step}
\end{figure}

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the process by computing new type argument tuples for a single formula. Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop. 
The separation of type argument tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some matchings processed in previous iterations. %Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
New tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one tuple component is instantiated by the substitution.

The total number of type argument tuples can increase cubicly in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth, for a single formula. If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type argument tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

Similar bounds are used for the non-monomorphic type argument tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.

The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}. Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
For two such symbol instances, types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the pseudocode, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant. The total number of substitutions generated is capped by \textcolor{ourblueviolet}{substitution\_limit}.

@@

   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\)}{
            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\)}{
               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
            }
            {\Break}

@@

}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for match generation}
\label{subst_gen}
\end{figure}

@@

   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
   \SetKwFunction{GenFormulae}{generate\_formulae}

@@

   \Return \NewFormulae
}
\end{algorithm}
%\end{quote}
\caption{Pseudocode for monomorphic formula generation}
\label{gen_formulae}
\end{figure}

The various bounds presented here overlap to some extent. For instance, having at most \textcolor{ourblueviolet}{substitution\_limit} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitution.

@@

\end{algorithm}
%\end{quote}
\caption{Pseudocode for monomorphising substitution generation}
\label{mono_substs}
\end{figure}

Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments. The last phase uses this set to instantiate the type variables in the input problem's non-monomorphic formuale. In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase. %As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.

The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
It generates monomorphising substitutions and applies them to the polymorphic formula of the input problem that they instantiate. Since the substitutions are monomorphising relative to the formula they applied to, the resulting formulae will be monomorphic. The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae. It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formulae it is given.

To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic. Such substitutions are computed using a recursive function. Given a set \(V\) of type variables and a set \(S(\varphi)\) of substitutions, it will select a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S(\varphi)\) that instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formula per initial polymorphic formula. Generating an excessive number of new formulae can flood the downstream prover. The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the algorithm's most explosive part. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
type variables, up to \(n^v\) monomorphising substitutions may be generated. Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

The monomorphisation algorithm is parametrised by many bounds. The first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition with and without E on polymorphic problems. The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first order and higher order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in at most 30 seconds with a single thread.
%This corresponds to the default time limit in Sledgehammer.

@@

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiations of type variables but takes more time. Ideally, all possible combinations of values for all bounds could be exhaustively tested to find the best compromise between completeness and speed. Given the number of bounds, however, this is unfeasible. Instead, we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

@@

Before finding values to assign to the bounds of the monomorphisation algorithm, we must make the choice of the base options used to run Zipperposition. The prover has a portfolio mode with several configurations. Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.

We conducted additional informal evaluations on the 500 problem suite to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
%
\begin{minipage}[t]{.5\textwidth}
\vskip0\topsep

>>>>

\begin{examplex}\rm
Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){.}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){.}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){.}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] 
         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){.} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

\looseness=-1
The algorithm presented above is too naive for an implementation. In this section, we present a lower level algorithm with the following features.
First, numeric bounds are introduced to stop explosive enumerations.
Second, type argument tuples are separated into an old set and a new set to avoid re-computing some of the same matchings in successive iterations.
Third, substitutions are directly applied to the type arguments instead of the formulae. This avoids having to re-extract the type arguments from the formulae at each iteration. New formulae are generated only once all iterations are complete, in a separate, final step.

The data structures used in the algorithm are based on the ones used in the high level description. Instead of a map \(M\) from symbols to monomorphic\pagebreak[2] type argument tuples, we now have \(M_\text{old}\) and \(M_\text{new}\), which play the same role whilst also distinguishing between those type argument tuples that have already been matched against and those that have not. Similarly, \(N_\text{old}\) and \(N_\text{new}\) replace the map \(N\) from formulae to symbols to non-monomorphic type argument tuples. Finally, maintain a map \(S\) from formulae to the substitutions generated by the matchings. It is used to generate new formulae in the final phase.

All sets referenced in the algorithm are %assumed to be
finite. Moreover, the algorithm relies on primitives whose implementation depends on the specifics of the grammar and logic used. %They will therefore not be expanded upon.
Functions computing the following are assumed to be available: 
\begin{enumerate}
   \item[\labelitemi] \emph{initialisation\((\Phi)\)}, where \(\Phi\) is a set of (polymorphic) formulae, extracts the initial type argument maps \(M\) and \(N\) from \(\Phi\).
   \item[\labelitemi] \emph{type\_variables\((\tau_1, \dots, \tau_n)\)}, where \(\tau_1, \dots,\tau_n\) are types, gathers all the type variables from each type \(\tau_1, \dots, \tau_n\) into a set.
   \item[\labelitemi] \emph{match\((\upsilon, \tau)\)}, where \(\upsilon\) and \(\tau\) are types, matches \(\upsilon\) against \(\tau\) and either fails or returns \emph{Success\((\sigma)\)}, where \(\sigma\) results from the matching. The algorithm matches only non-monomorphic types against monomorphic types.
   \item[\labelitemi] \emph{domain\((\sigma)\)} %, where \(\sigma\) is a substitution,
   returns the set of type variables \(\alpha\) such that \(\sigma(
   \alpha) \not= \alpha\).
   \item[\labelitemi] \emph{compatible\((\sigma_1, \sigma_2)\)} %, where \(\sigma_1\) and \(\sigma_2\) are substitutions,
   tests the compatibility between \(\sigma_1\) and \(\sigma_2\).
%   \item[\labelitemi] Composition and application of substitutions are written using mathematical notations.
   \item[\labelitemi] \emph{mangle\((\Phi)\)}, where \(\Phi\) is a set of monomorphic formulae, returns the same set of formulae where all types have been mangled.
\end{enumerate}

@@

   \Return \(\Mangle(\GenFormulae(\Phi, S))\)
}
\end{algorithm}
%\end{quote}
\caption{Algorithm for iterative monomorphisation}
\label{iter_mono}
\end{figure}

The iterative monomorphisation algorithm is given in Figure~\ref{iter_mono}. It has three phases. The first phase applies a \emph{monomorphisation step} to each formula in \(\Phi\) until the user-set limit, \textcolor{ourblueviolet}{num\_loops}, is reached. This limit is the only bound necessary for the algorithm to terminate. We use the colour blue to identify bounds and code related to bounds. At the end of each of these iterations, the old and new type argument maps are updated with newly generated types. Once these iterations are completed, the first phase is done and the substitutions used to create new type argument tuples are passed to $\mathit{generate\_mono\_formulae}$ for the second phase. The third phase mangles the composite types of the newly monomorphised formulae. This allows targeting a simply typed logic with no support for $n$-ary type constructors.
%~\cite[Sections 4 and 5]{mono-trans}.
% Paper name is gt

@@

   \SetKw{Let}{let}

@@

   \Let $S' = \SubstGen(M_{\text{new}}, N_{\text{new}}(\varphi))
   \cup \SubstGen(M_{\text{new}}, N_{\text{old}}(\varphi))
   \cup \SubstGen(M_{\text{old}}, N_{\text{new}}(\varphi))$\;
   \ForEach{\(\sigma\in S'\)}{
   \ForEach{\((f\mapsto (\upsilon_1, \dots,\upsilon_n))\in N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi)\)}{
            \eIf{\(\TyVars(\upsilon_1, \dots, \upsilon_n) \subseteq \Domain(\sigma)\)}{
               \textcolor{ourblueviolet}{
               \uIf{\(|M_{\text{next}}| < \MaxMono\)}{
                  \textcolor{black}{
                  \(M_{\text{next}}(f)\leftarrow M_{\text{next}}(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\kern-300mm\}\)\;
                  \UsedSubst \(\leftarrow \UsedSubst \cup\{\sigma\}\)\;}
               }
               \uElseIf{\(|N_{\text{next}}(\varphi)| \geq \MaxPoly\) }{
                  \(M_{\text{next}} \leftarrow M_{\text{next}}\setminus (M_{\text{old}} \cup M_{\text{new}})\)\;
                  \(N_{\text{next}}(\varphi)\leftarrow N_{\text{next}}(\varphi)\setminus (N_{\text{old}}(\varphi)\cup N_{\text{new}}(\varphi))\kern-300mm\)\;
                  \Return \((M_{\text{next}}, N_{\text{next}}(\varphi), \UsedSubst)\)
               }}
            }{\textcolor{ourblueviolet}{
               \If{\(|N_{\text{next}}(\varphi)| < \MaxPoly\)}{ \textcolor{black} {
               \(N_{\text{next}}(\varphi)(f)\leftarrow N_{\text{next}}(\varphi)(f)\cup\{(\sigma(\upsilon_1),\dots,\sigma(\upsilon_n))\}\kern-300mm\)\;
               \(\UsedSubst \leftarrow \UsedSubst \cup\{\sigma\}\)\;
               }}
            }}
      }
   }

@@

\end{algorithm}
%\end{quote}
\caption{Algorithm for formula monomorphisation step}
\label{mono_step}
\end{figure}

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the process by computing new type argument tuples for a single formula. Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop. 
The separation of type argument tuples into old and new maps is used to ensure that only combinations involving at least one new map are considered. This avoids re-computing some matchings processed in previous iterations. %Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
New tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one tuple component is instantiated by the substitution.

The total number of type argument tuples can increase cubically in the number of type argument tuples at each iteration and can therefore grow doubly exponentially in the number of iterations. We sketch an example of such growth, for a single formula. If we assume that after \(k\) iterations, there are \(N_k\) total type argument tuples divided evenly between monomorphic and non-monomorphic type argument tuples, then there can be up to \(\bigl(\frac{N_k}{2}\bigr)^2\) successful matches, yielding as many substitutions. Each substitution is then applied to each non-monomorphic type argument for a total of \(\bigl(\frac{N_k}{2}\bigr)^3\) possible new type argument tuples. If half of these new type argument tuples are monomorphic and the other half are non-monomorphic, then the next iteration will begin with \(N_{k+1} = N_k^3 \cdot 2^{-3}\) evenly split type argument tuples. Therefore, the total number of type argument tuples on the \(k\)th iteration can reach \(N_k = N_0^{3^k} \cdot 2^{\frac{-3^{k+2}+3}{2}}\).

@@

Similar bounds are used for the non-monomorphic type argument tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and iteration.

The \emph{matches} function, which computes the substitutions used for generating new type arguments, is given in Figure~\ref{subst_gen}. Each symbol instance from \(N(\varphi)\) is matched against all corresponding symbol instances from \(M\).
For two such symbol instances, types from the non-monomorphic type argument tuple are matched component-wise against the types from the monomorphic type argument tuple. The resulting substitutions are composed if they are compatible. In the algorithm, this is checked by making sure the \textbf{while} loop has successfully iterated over all elements of the type argument tuple. If any substitutions are incompatible, the matchings are discarded. Since the composition of two compatible substitutions is commutative, the order of composition is irrelevant. The total number of substitutions generated is limited by \textcolor{ourblueviolet}{substitution\_limit}.

@@

   \ForEach{\(f\mapsto (\upsilon_1, \dots,\upsilon_n)\in N(\varphi)\)}{
      \ForEach{\( (\tau_1, \dots,\tau_n) \in M(f) \)}{
         \(\sigma\leftarrow \{\}\)\;
         \(i \leftarrow 1\)\;
         \While{\(i \leq n\)}{
            \eIf{\(\Match(\upsilon_i, \tau_i)\) has the form \(\Success(\sigma_i)\) \And \(\Compatible(\sigma, \sigma_i)\kern-300mm\)}{
               \(\strut\sigma\leftarrow\sigma_i\circ\sigma\)
            }
            {\Break}

@@

}
\end{algorithm}
%\end{quote}
\caption{Algorithm for match generation}
\label{subst_gen}
\end{figure}

%%% @TANGUY: Jannis wrote:
% "I feel like this pseudocode could be simplified a lot by taking a
% higher-level view of the while loop. E.g.

% if for all v_i and t_i, match(v_i, t_i) = Success(σ_i) and the σ_i are all 
% compatible then
%    σ := σ_1 \comp ... \comp σ_n

% The fact that you break early is imo trivial enough that it doesn't need to be
% discussed explicitly."

@@

%%% @TANGUY: Jannis wrote:
% "All the other limits are named something like max_x or x_cap."
   \SetKwFunction{Init}{initialisation}\SetKwFunction{FStep}{formula\_mono\_step} 
   \SetKwFunction{TyVars}{type\_variables}\SetKwFunction{MonoSubst}{mono\_substs}
   \SetKwFunction{GenFormulae}{generate\_formulae}

@@

   \Return \NewFormulae
}
\end{algorithm}
%\end{quote}
\caption{Algorithm for monomorphic formula generation}
\label{gen_formulae}
\end{figure}

The various bounds presented here overlap to some extent. For instance, having at most \textcolor{ourblueviolet}{substitution\_limit} substitutions generated by \emph{matches} may be sufficient to curb the number of new type argument tuples, making the \textcolor{ourblueviolet}{\MonoCap}, \textcolor{ourblueviolet}{\MonoMult}, \textcolor{ourblueviolet}{\MonoFloor} triplet superfluous. Each bound nonetheless has its uses. For example, problems that lead to few successful matches but many type argument tuples may benefit from a limit on the number of new type argument tuples whilst problems for which substitution generation is more explosive may benefit from a limit on the number of generated substitutions.

@@

\end{algorithm}
%\end{quote}
\caption{Algorithm for monomorphising substitution generation}
\label{mono_substs}
\end{figure}

Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments. The last phase uses this set to instantiate the type variables in the input problem's non-monomorphic formulae. In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase. %As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.

The \emph{generate\_mono\_formulae} function is given in Figure~\ref{gen_formulae}.
It generates monomorphising substitutions and applies them to the polymorphic formulae of the input problem that they instantiate.
A substitution \(\sigma\) is \emph{monomorphising} for a formula \(\varphi\) if \(\sigma(\varphi)\) is monomorphic.
Since the substitutions are monomorphising relative to the formula they are applied to, the resulting formulae will be monomorphic. The \textcolor{ourblueviolet}{total\_new\_formulae} bound is used to control the total number of new formulae. It overlaps with \textcolor{ourblueviolet}{max\_substs} but can be useful to set an absolute limit on the size of the final problem.
%In the Zipperposition implementation, this bound is one of the more important ones as the performance of the E prover can be significantly affected by the number of formulae it is given.

To monomorphise a polymorphic formula, we first compute its {monomorphising substitutions} using the \emph{mono\_substs} function given in Figure~\ref{mono_substs}.
Such substitutions are computed using a recursive function. Given a set \(V\) of type variables and a set \(S(\varphi)\) of substitutions, it selects a substitution \(\sigma_{\negvthinspace\Delta}\) from \(S(\varphi)\) that instantiates at least one of the type variables in \(V\). It is important that \(\sigma_{\negvthinspace\Delta}\) be compatible with \(\sigma\) so that they can be composed and the function recursively called to instantiate the remaining type variables.

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formulae per initial polymorphic formula. Generating an excessive number of new formulae can overwhelm the downstream prover. The final number of output formulae is limited to at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the algorithm's most explosive part. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
type variables, up to \(n^v\) monomorphising substitutions may be generated. Recall that the total number of type argument tuples used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
%%% @TANGUY: About the last sentence, Jannis writes:
%%% "Not immediately clear to me what this entails."
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

The monomorphisation algorithm is parametrised by many bounds. The first part of the evaluation process seeks appropriate values for these bounds. The second part compares the performance of Zipperposition without E and with the new monormorphising E backend on polymorphic problems. The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first order and higher order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems
that could be solved by the prover in at most 30 seconds per problem with a single thread.

%%% @TANGUY: Jannis writes:
% "Also, does the 30-sec timeout include the monomorphisation pass itself?"

%This corresponds to the default time limit in Sledgehammer.

@@

Each bound of the monomorphisation process presents a tradeoff: a higher bound allows for a more exhaustive instantiation of type variables but takes more time. Since we cannot test all possible combinations of values for all bounds to find the best compromise between completeness and speed,
we group closely related bounds together and test combinations of values for the bounds in these groups. Once we find the best performing set of values for a group, we assign these values to the corresponding bounds as we begin the search for the next group. If several groups result in the same number of proved problems, we select the most constraining values. Winning entries are shown in bold in Tables \ref{mono_ty_args} to \ref{e_settings}.

@@

Before finding values to assign to the bounds of the monomorphisation algorithm, we must choose which base options to run Zipperposition with. The prover has a portfolio mode with several configurations. Because the space of possible base configurations is too large to evaluate exhaustively, we evaluated, in a preliminary experiment, all portfolio configurations that called E against our benchmark suite of 500 problems, but we disabled E.
Our preliminary evaluation found that the so-called \verb|40_b.comb| configuration performed best by proving 131 problems.
This configuration became the base configuration, which we used as a basis to evaluate the monomorphisation options.

We conducted additional informal evaluations on the 500 problems to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
%
\begin{minipage}[t]{.5\textwidth}
\vskip0\topsep
